---
title: "Exercise 2"
author: "Mads Adrian Simonsen, William Scott Grundeland Olsen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  github_document: default
  html_document:
    df_print: paged
subtitle: TMA4300 Computer Intensive Statistical Models
---

```{r setup, include=FALSE}
library(rmarkdown)  # Dynamic Documents for R
library(knitr)  # A General-Purpose Package for Dynamic Report Generation in R
knitr::opts_chunk$set(echo = TRUE, tidy = FALSE, message = FALSE, warning = FALSE,
                      strip.white = TRUE, prompt = FALSE, cache = TRUE,
                      size = "scriptsize", fig.width = 6, fig.height = 4)
```

\newcommand{\E}{\operatorname E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\Corr}{\operatorname{Corr}}
\newcommand{\Exp}{\operatorname{Exponential}}
\newcommand{\Uniform}{\operatorname{Uniform}}
\newcommand{\Betadist}{\operatorname{Beta}}
\newcommand{\Normal}{\operatorname{Normal}}
\newcommand{\SD}{\operatorname{SD}}
\newcommand{\RSS}{\mathrm{RSS}}
\newcommand{\MSE}{\mathrm{MSE}}
\newcommand{\T}{\mathsf T}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\Bias}{\operatorname{Bias}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\argmax}{\operatorname*{arg\,max}}
\newcommand{\bfy}{\mathbf{y}}
\newcommand{\e}{\mathrm{e}}


```{r libraries, eval = TRUE, echo = FALSE}
library(tidyverse)  # Collection of R packages designed for data science

library(INLA)       # Full Bayesian Analysis of Latent Gaussian Models using Integrated
                    # Nested Laplace Approximations
```

# Problem A: The coal-mining disaster data

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 3.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 4.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 5.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 6.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 7.

### Subsubproblem (a)

### Subsubproblem (b)


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 8.


[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------


# Problem B: INLA for Gaussian Data

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 3.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 4.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 5.


[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------