---
title: "Exercise 3"
author: "Mads Adrian Simonsen, William Scott Grundeland Olsen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  github_document: default
  html_document:
    df_print: paged
subtitle: TMA4300 Computer Intensive Statistical Models
urlcolor: blue
bibliography: ref.bib
---

```{r setup, include=FALSE}
library(rmarkdown)  # Dynamic Documents for R
library(knitr)  # A General-Purpose Package for Dynamic Report Generation in R
knitr::opts_chunk$set(
  echo = TRUE, tidy = FALSE, message = FALSE, warning = FALSE, strip.white = TRUE,
  prompt = FALSE, cache = TRUE, size = "scriptsize", fig.width = 6, fig.height = 4
)
knitr::opts_knit$set(root.dir = "./Additional files")  # Changed working directory
```

\newcommand{\E}{\operatorname E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\Corr}{\operatorname{Corr}}
\newcommand{\Poisson}{\operatorname{Poisson}}
\newcommand{\Exp}{\operatorname{Exponential}}
\newcommand{\Uniform}{\operatorname{Uniform}}
\newcommand{\Betadist}{\operatorname{Beta}}
\newcommand{\Gammadist}{\operatorname{Gamma}}
\newcommand{\InvGamma}{\operatorname{Inv-Gamma}}
\newcommand{\Normal}{\operatorname{Normal}}
\newcommand{\SD}{\operatorname{SD}}
\newcommand{\RSS}{\mathrm{RSS}}
\newcommand{\MSE}{\mathrm{MSE}}
\newcommand{\T}{\mathsf T}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\Bias}{\operatorname{Bias}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\argmax}{\operatorname*{arg\,max}}
\newcommand{\bfy}{\mathbf{y}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bfzero}{\mathbf{0}}
\newcommand{\bfeta}{\boldsymbol{\eta}}
\newcommand{\dtheta}{\, \mathrm{d} \theta}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\e}{\mathrm{e}}


```{r libraries and help files, eval = TRUE, echo = FALSE}
library(tidyverse)  # Collection of R packages designed for data science

# Extract pre-programmed R-code
source("probAdata.R")  # Code
source("probAhelp.R")  # Data
```

# Problem A: Comparing AR(2) parameter estimators using resampling of residuals

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.


[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------

# Problem B: Permutation test

[Bilirubin](http://en.wikipedia.org/wiki/Bilirubin) is a breakdown product of hemoglobin, which is a principal component of red blood cells. If the liver has suffered degeneration,  the decomposition of hemoglobin is elevated, or the gall bladder has been destroyed, large amounts of bilirubin can accumulate in the blood, leading to jaundice, which is a yellowish or greenish pigmentation in the skin.

We will look at a data set taken from [@jorgensen1993theory]. It contains measurements of the concentration of bilirubin (mg/dL) in blood samples taken from an three young men.

## Subproblem 1.
```{r echo = FALSE}
bilirubin <- read.table("bilirubin.txt", header = TRUE)
bilirubin %>%
  group_by(pers) %>% 
  mutate(row = row_number()) %>% 
  pivot_wider(names_from = pers, values_from = meas) %>% 
  select(-row) %>% 
  t() %>%
  kable(booktabs = TRUE) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::add_header_above(c("Individual", "Concentration (mg/dL)" = 11))
```


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.

[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 3.

[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------

# Problem C: The EM-algorithm and bootstrapping

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.

[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 3.

[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 4.

[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------

# References