---
title: "Exercise 3"
author: "Mads Adrian Simonsen, William Scott Grundeland Olsen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  github_document: default
  html_document:
    df_print: paged
subtitle: TMA4300 Computer Intensive Statistical Models
---

```{r setup, include=FALSE}
library(rmarkdown)  # Dynamic Documents for R
library(knitr)  # A General-Purpose Package for Dynamic Report Generation in R
knitr::opts_chunk$set(
  echo = TRUE, tidy = FALSE, message = FALSE, warning = FALSE, strip.white = TRUE,
  prompt = FALSE, cache = TRUE, size = "scriptsize", fig.width = 6, fig.height = 4
)
knitr::opts_knit$set(root.dir = "./Additional files")  # Changed working directory
```

\newcommand{\E}{\operatorname E}
\newcommand{\Var}{\operatorname{Var}}
\newcommand{\Cov}{\operatorname{Cov}}
\newcommand{\Corr}{\operatorname{Corr}}
\newcommand{\Poisson}{\operatorname{Poisson}}
\newcommand{\Exp}{\operatorname{Exponential}}
\newcommand{\Uniform}{\operatorname{Uniform}}
\newcommand{\Betadist}{\operatorname{Beta}}
\newcommand{\Gammadist}{\operatorname{Gamma}}
\newcommand{\InvGamma}{\operatorname{Inv-Gamma}}
\newcommand{\Normal}{\operatorname{Normal}}
\newcommand{\SD}{\operatorname{SD}}
\newcommand{\RSS}{\mathrm{RSS}}
\newcommand{\MSE}{\mathrm{MSE}}
\newcommand{\T}{\mathsf T}
\newcommand{\tr}{\operatorname{tr}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\Bias}{\operatorname{Bias}}
\newcommand{\diag}{\operatorname{diag}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\argmax}{\operatorname*{arg\,max}}
\newcommand{\bfy}{\mathbf{y}}
\newcommand{\bff}{\mathbf{f}}
\newcommand{\bfzero}{\mathbf{0}}
\newcommand{\bfeta}{\boldsymbol{\eta}}
\newcommand{\dtheta}{\, \mathrm{d} \theta}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\e}{\mathrm{e}}


```{r libraries and help files, eval = TRUE, echo = FALSE}
library(tidyverse)  # Collection of R packages designed for data science

# Extract pre-programmed R-code
source("probAdata.R")  # Code
source("probAhelp.R")  # Data
```

# Problem A: Comparing AR(2) parameter estimators using resampling of residuals

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.


[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------

# Problem B: Permutation test

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.

[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 3.

[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------

# Problem C: The EM-algorithm and bootstrapping

## Subproblem 1.


[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 2.

[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 3.

[//]: # ----------------------------------------------------------------------------------------------------------------

## Subproblem 4.

[//]: # ----------------------------------------------------------------------------------------------------------------
[//]: # ----------------------------------------------------------------------------------------------------------------